---
title: "Exploring Socioeconomic Patterns in the 2021 Household Census"
author: "Oluwadamilola Oreofe Ajakaiye"
format:
  html:
    code-fold: true
    code-link: true
    toc: true
---

```{r}
# | label: loading packages
# | include: false
library(tidyverse)
library(ggplot2)
library(ggthemes)
```

```{r}
#| label: Importing the Data
#| include: false
data <- read_csv("data.csv")
```

# 1.0 Introduction

This report explores the relationships between demographic factors (age, marital status) and socioeconomic variables (income, education level) within a 2021 household census dataset.

## 1.1 Data Preprocessing

Check for missing values in important columns

```{r}
# View the first few rows of the data
head(data)
```

```{r}
# Check for missing values in important columns
sum(is.na(data$Age))          # Missing values in Age column
sum(is.na(data$Mar_Stat))     # Missing values in Marital Status column
sum(is.na(data$INC))          # Missing values in Income column
sum(is.na(data$Eth))          # Missing values in Ethnicity column

```

The results from checking for missing values in important columns show the following:

1.  **Age** column: `[1] 0` - No missing values.

2.  **Marital Status (Mar_Stat)** column: `[1] 6144` - 6,144 missing values.

3.  **Income (INC)** column: `[1] 6173` - 6,173 missing values.

4.  **Ethnicity (Eth)** column: `[1] 0` - No missing values.

Next, Missing values are inputed

```{r}
# Impute missing values in the Income (INC) column with the median
data$INC[is.na(data$INC)] <- median(data$INC, na.rm = TRUE)

# Check if there are any remaining missing values in the Income column
sum(is.na(data$INC))  # Should return 0

```

Function to calculate mode

```{r}
# Function to calculate mode
mode_function <- function(x) {
  unique_x <- unique(x)
  unique_x[which.max(tabulate(match(x, unique_x)))]
}

# Impute missing values in the Marital Status (Mar_Stat) column with the mode
data$Mar_Stat[is.na(data$Mar_Stat)] <- mode_function(data$Mar_Stat)

# Check if there are any remaining missing values in the Marital Status column
sum(is.na(data$Mar_Stat))  # Should return 0

```

Drop unnecessary columns

```{r}
# Drop unnecessary columns
data_cleaned <- data[, !(names(data) %in% c("ID", "Person_ID"))]

# Check the structure of the cleaned dataset
str(data_cleaned)


```

Convert variables to factors

```{r}
# Convert variables to factors
data_cleaned$Female <- as.factor(data_cleaned$Female)
data_cleaned$Mar_Stat <- as.factor(data_cleaned$Mar_Stat)
data_cleaned$Eth <- as.factor(data_cleaned$Eth)
data_cleaned$`Highest Ed` <- as.factor(data_cleaned$`Highest Ed`)
```

Create age ranges

```{r}
data_cleaned <- data_cleaned %>%
  mutate(Age_Range = case_when(
    Age < 18 ~ "<18",
    Age >= 18 & Age < 30 ~ "18-29",
    Age >= 30 & Age < 40 ~ "30-39",
    Age >= 40 & Age < 50 ~ "40-49",
    Age >= 50 & Age < 60 ~ "50-59",
    Age >= 60 ~ "60+",
  ))
```

## 2.0 Visualisations

Boxplot of Income by Marital Status

```{r}
# Boxplot of Income by Marital Status
ggplot(data_cleaned, aes(x = Mar_Stat, y = INC, fill = Mar_Stat)) +
  geom_boxplot() +
  labs(title = "Income Distribution by Marital Status", x = "Marital Status", y = "Annual Income (£)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Boxplot of Income by Age Ranges

```{r}
# Create a boxplot of Income by Age Ranges
ggplot(data_cleaned, aes(x = Age_Range, y = INC, fill = Age_Range)) +
  geom_boxplot() +
  labs(title = "Income Distribution by Age Ranges", x = "Age Range", y = "Annual Income (£)") +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend title

```

Bar plot of Highest Education Level by Ethnicity

```{r}
# Bar plot of Highest Education Level by Ethnicity
ggplot(data_cleaned, aes(x = Eth, fill = `Highest Ed`)) +
  geom_bar(position = "fill") +  # Stacked bar chart to show proportions
  labs(title = "Proportion of Highest Education Level by Ethnicity", x = "Ethnicity", y = "Proportion") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +  # You can change the palette if necessary
  theme(legend.title = element_blank())  # Remove legend title
```

## 3.0 Interesting Things Discovered

1.  Box plot of Income by Marital Status: Married individuals shows a higher median income compared to those with other marital statuses, suggesting that marital status may influence financial stability.

2.  Box plot of Income by Age Ranges: The age range of 30 to 39 appears to have the highest average income. However, individuals aged 60 and above tend to have more income earners within that group. In contrast, the 18 and over age group has the fewest income earners and records the lowest income levels.

3.  Bar Plot of Highest Education Level by Ethnicity: The plot indicates that individuals of Hispanic descent have the highest proportion of higher education attainment

## 4.0 Conclusion

In summary, the analysis reveals key insights into how demographic factors relate to income and education levels. Married individuals tend to have higher median incomes, indicating a potential link between marital status and financial stability. Age also plays a significant role, with those aged 30 to 39 having the highest average income, while older age groups have more income earners. Additionally, ethnicity appears to influence educational attainment, with individuals of Hispanic descent showing the highest proportion of higher education. These findings highlight the interplay between marital status, age, and ethnicity in shaping income and educational outcomes.

Author: Oluwadamilola Oreofe Ajakaiye

Student ID: 2439749

Program: MSc. Data Science

Module: 7CS034 Data Science
